---
title: "MA132 Enrollment Analysis"
author: "Peter-John King"
date: "21 Sep 2020"
output: html_notebook
---

The intention of this analysis is to predict the number of students who will be enrolled 
in MA132 for Spring 2021

## Load packages

```{r}
library(tidyverse)
library(janitor)
library(readxl)
library(skimr)
library(summarytools)
```


## Read and clean the data
Loads the enrollment data from the excel spreadsheet

```{r}
dfa <- read_xlsx("data/CU_ACAD_DEPT_ENRL.xlsx") %>% clean_names()
```

```{r}

skim(dfa)
```

### Tidy Data

Let us separate column term 2 into two variables 
```{r}

dfa <- dfa %>% separate(term2, c("semester", "year"), " ")

```

### Clean and Subset Data

We get rid of the following variables:

* acad_org
* current_status
* professor
* course_id

Filtering to only include:
* math courses
* catalog number 131 and 132
* lectures

We make the year numeric 

We remove duplicate data 

```{r}

dfb <- dfa %>% select(subject:component) %>% 
  filter(subject == "MA") %>% filter (catalog %in% c("131", "132")) %>%
  filter(component == "LEC") %>% select(-component) %>% distinct() %>%
  mutate(year=as.numeric(year))
```

Now, we find the total students enrolled in each smester
```{r}
dfc <- dfb %>% group_by(catalog, semester, year) %>%
  summarize(tot_enrl = sum(tot_enrl)) %>%
  filter(year!=2021)


```

In order to compare using a scatterplot, I will need to make a "wide" dataframe.

Additional, We want to consider the "previous" fall, so we add a column to the dataframe which is lagged.  

We remove 2021, as there is no valid data.

```{r}
dfd <- dfc %>% pivot_wider(values_from = tot_enrl,names_from=c(semester,catalog)) %>% 
  clean_names() %>%
  mutate(prev_131_fall=lag(fall_131),prev_132_fall=lag(fall_132)) %>% 
  filter(year!=2021)
```

We now plot fall ma 131 with spring ma 132 to see if a correlation exists. We found that with the exception of 2014, we found that a linear correlation existed. 


```{r}
dfd %>% ggplot(aes(x=prev_131_fall,y=spring_132)) +geom_point()
```
